Visual Diagnostics for Constrained Optimisation with Application to Guided Tours

### 2021-09-21 paper/zhang-cook-laa-langrene-menendez.Rmd
response to review revised

Several annotations have been made to help understanding this plot. As mentioned previously, the original basis space is a high-dimensional sphere and random bases on the sphere can be generated via the \CRANpkg{geozoo} [@geozoo] package. We use PCA to estimate the 2D space that is best suited to visualise the bases. The centre of the 2D view is the first two PC coordinates of the zero matrix, and the radius is estimated as the largest distance from the centre to the basis. The theoretical best basis is known for simulated data and can be labelled to compare how close the end basis found by the optimisers is to the theoretical best. Various aesthetics, i.e. size, alpha (transparency), and colour, are applicable to emphasize critical elements and adjust for the presentation. For example, anchor points and search points are less important and hence a smaller size and alpha are used. Alpha can also be applied on the interpolation paths to show the start to finish from transparent to opaque.

@@

```{r toy-pca, fig.height = 4, fig.asp = 1, fig.cap = "Search paths of CRS (green) and PD (brown) in the PCA-reduced basis space for 1D projection problem on the five-variable dataset, \\code{boa5} using holes index. The basis space, a 5D unit sphere, is projected onto a 2D circle by PCA. The black star represents the theoretical best basis the optimisers aiming to find. All the bases in PD have been flipped for easier comparison of the final bases and a grey dashed line has been annotated to indicate the symmetry of the two start bases."}
```

@@

One interesting situation observed in the previous examples is that, for some simulations, as shown on the left panel of Figure \ref{fig:flip-sign}, the target basis is generated on the other half of the basis space and the interpolator seems to draw a straight line to interpolate. Bases with opposite signs do not affect the projection and index value, but we would prefer the target to have the same orientation as the current basis. The orientation of two bases can be checked via calculating the determinant and a negative value suggests the two bases having different orientation. For 1D bases, this can be corrected by flipping the sign in one of the basis while the correction for a general p-D bases is more complex and beyond the scope of this work. Here an orientation check is carried out once a new target basis is generated and the sign in the target basis will be flipped if a negative determinant is obtained. The interpolation after implementing the orientation check is shown on the right panel of Figure \ref{fig:flip-sign} where the unsatisfactory interpolation no longer exists.

@@

This paper has provided several visual diagnostics that can be used for understanding a complex optimisation procedure and are implemented in the ferrn package. The methods were illustrated using the optimisers available for projection pursuit guided tour. Here the constraint is the orthonormality condition of the projection bases. The approach used broadly applies to other constrained optimisers, by saving the path of the optimiser in a form required to input into the ferrn package. These high-dimensional data objects can then be viewed using the tour.

The progressive optimisation of a target function and it's coverage of the search space can be viewed in both reduced 2D space and the full space. These visualisation can lead to insights for evaluating and comparing the performance of multiple optimisers in a specific problem. They can provide better understanding of existing methods or motivate the development of new approaches. For example, we have compared how three optimisers perform when maximising a non-smooth index function and illustrated how the pseudo-derivative search fails in this setting. The observations from our experiments have also been translated into improved optimisation methods for the guided tour, e.g. we introduced the option to interrupt the search if a better basis is found along the path.

>>>>

Several annotations have been made to help understanding this plot. As mentioned previously, the original basis space is a high-dimensional sphere and random bases on the sphere can be generated via the \CRANpkg{geozoo} [@geozoo] package. We use PCA to project and visualize the parameters/bases in 2D. The centre of the 2D view is the first two PCs of the data matrix. It theoretically should be a circle, but may have some irregular edges due to finite sampling. Thus the edge is smoothed by using a radius estimated as the largest distance from the centre to any basis. In the simulation, the theoretical best basis is known and can be labelled to compare how close to this that the optimisers stopped. Various aesthetics, i.e. size, alpha (transparency), and colour, are applicable to emphasize critical elements and adjust for the presentation. For example, anchor points and search points are less important and hence a smaller size and alpha are used. Alpha can also be applied on the interpolation paths to show the start to finish from transparent to opaque.

@@

```{r toy-pca, fig.height = 4, fig.asp = 1, fig.cap = "Search paths of CRS (green) and PD (brown) in the PCA-reduced basis space for 1D projection problem on the five-variable dataset, \\code{boa5} using holes index. The basis space, a 5D unit sphere, is projected onto a 2D circle by PCA. The theoretical best basis is indicated by the black star. All the bases in PD have been flipped for easier comparison of the final bases and a grey dashed line has been annotated to indicate the symmetry of the two start bases."}
```

@@

One interesting situation observed in the previous examples is that, for some simulations, as shown on the left panel of Figure \ref{fig:flip-sign}, the target basis is generated on the other half of the basis space and the interpolator seems to draw a straight line to interpolate. Bases with opposite signs do not affect the projection and index value, but we would prefer the target to have the same orientation as the current basis. The orientation of two bases can be computationally checked by calculating the determinant -- a negative value suggests the two bases having different orientation. For 1D bases, this can be corrected by flipping the sign in one basis. For higher dimensions, it can be a bit more difficult because the orthonormality of the basis needs to be maintained also when an individual vector is flipped. Here an orientation check is carried out once a new target basis is generated and the sign in the target basis will be flipped if a negative determinant is obtained. The interpolation after implementing the orientation check is shown on the right panel of Figure \ref{fig:flip-sign} where the unsatisfactory interpolation no longer exists.

@@

This paper has provided several visual diagnostics that can be used for understanding a complex optimisation procedure and are implemented in the ferrn package. The methods were illustrated using the optimisers available for projection pursuit guided tour. Here the constraint is the orthonormality condition of the projection bases, which corresponds to optimisation over spheres and torii. The approach described broadly applies to other constrained optimisers. Although the manifold in $p$-space might be different the diagnostic techniques are the same. A researcher would begin by saving the path of the optimiser in a form required to input into the ferrn package, as described in this paper. One might generally make more samples from the constrained space upon which to assess and compare the optimisation paths. These high-dimensional data objects can then be viewed using the tour.

The progressive optimisation of a target function and it's coverage of the search space can be viewed in both reduced 2D space and the full space. These visualisations can lead to insights for evaluating and comparing the performance of multiple optimisers operating on the same task. They can provide better understanding of existing methods or motivate the development of new approaches. For example, we have compared how three optimisers perform when maximising a non-smooth index function and illustrated how the pseudo-derivative search fails in this setting. The observations from our experiments have also been translated into improved optimisation methods for the guided tour, e.g. we introduced the option to interrupt the search if a better basis is found along the path.
